Using a Highly Dynamic Language for Development

### 2021-10-16 gic2021-highly-dynamic/paper.tex
Update stuff with feedback from Robert

We explore some of the aspects of Common Lisp that make it particularly suited for games in detail, and also discuss some of the pitfalls we encountered and how to combat them.

@@

In CLOS methods are not attached to classes, but are instead parts of a generic function. Methods ``specialise'' on one or several of the required arguments of the function to a class. When the generic function is called, the system must first determine the set of ``applicable methods''. This set depends on the method combination used by the generic function. For brevity's sake, we will only look at the standard method combination here.

@@

As a brief example, let us consider \autoref{lst:method definition}. Here we define a generic function of two arguments called \code{handle}, as well as three methods. All of the methods require the first argument to be a subclass of \code{tick}. Method 1 and 2 also require the second argument to be a subclass of \code{player}, and method 3 a subclass of \code{enemy}.

@@

This method combination mechanism only really shines once we consider inheritance and especially multiple inheritance and the arising ``mixin classes''. Let us now define the classes we used for the previous listing. \autoref{lst:inheritance} shows five classes being defined, with \code{tick} being a subclass of \code{event}, and \code{player} and \code{enemy} being subclasses of \\\code{physics-object}.

@@

In other OOP paradigms a similar behaviour can usually be achieved by calling the superclass' method with \code{super}, but notice here that the behaviour of method 2 does not require changing anything about the other methods or subclasses. It thus remains wholly encapsulated in its own behaviour.

So far so good. Now imagine that we decide enemies, should emit a light so that they're always visible. To implement this, we're going to define a new class, \code{emitter}, and define the light flicker behaviour in a new \code{handle} method as shown in \autoref{lst:emitter}.

@@

To make the enemy adopt this new behaviour, all we need to do is add \code{emitter} to the \code{enemy} class' superclass list. Thanks to the method combination, calls to \code{handle} will now include the \code{emitter}'s method, and we have achieved the combination of the two behaviours.

@@

The order of the superclasses here is important. Classes that appear earlier in the list have ``precedence'' and are thus considered to be ``more specific'' when determining the set of applicable methods. By having \code{emitter} appear after \code{physics-object}, we ensure that the \code{emitter}'s \code{:before} method runs \textit{after} that of \code{physics-object}, ensuring the collisions have already resolved properly once we consider the lighting update.

@@

One issue that crops up when segregating behaviours into such small compartments is that you might not have enough control over the combination of them. The standard method combination offers no fine-grained control to exclude or reorder methods for a specific specialisation of an argument. One could devise their own method combination strategy to allow this kind of specialised behaviour, however we are not convinced that this would not lead to an ultimately even more confusing design.

While the standard method combination and mixins offer a great deal of flexibility out of the box, great care must still be taken when designing the overall class hierarchy and function protocol. Otherwise behaviours will not combine cleanly and lead to strange bugs, or might not combine properly at all.

@@

The first part is what's called ``conditions'', or Exceptions in other languages. Conditions are ``signalled'' (thrown) and can then be ``handled'' (caught) by a piece of code lower down in the stack. When handled, the stack unwinds to the point of the handler, and the matching handler function is invoked to resolve the issue. For our purposes here, Common Lisp offers two types of conditions: warnings and errors. When a warning is signalled that is not handled further up the stack, it simply vanishes and execution continues from the signalling point unimpeded. If an error remains unhandled, a dynamic debugger is invoked instead and the signalling thread is paused. From there one of the unique features comes to shine.

@@

This means that unlike traditional languages where an error causes a stack trace and then tries to continue or simply crashes, in Common Lisp execution is halted, waiting for you to analyse and potentially fix the bug. Once you're confident you know what to do or have solved the issue, you can pick from one of several ways to continue the execution from the error. This leaves the program running at all times and means you don't need to worry about long recompilation times or setup times to reproduce the error, you can immediately continue where you left off.

@@

The final piece is the existence of \code{handler-bind}. This is a type of handler for conditions, but unlike the regular handlers that cause the stack to unwind, these handlers are invoked \textit{on top of the stack} of where the condition was signalled. As such they have full access to the dynamic environment at error point. In fact, this is also how the top level debugger is invoked. However, with \code{handler-bind} you can automate the error resolution and avoid requiring user input.

A system with well-designed recovery points through restarts can then be set up to automatically recover from most errors in a way much more refined than traditional Exceptions, as there is an inversion of control.

@@

However, importantly, for the \code{main} \code{render}\\ method, this restart is automatically invoked, ensuring that any potential crashes at runtime of the game are gracefully ignored. This handler could also be defined elsewhere even higher up the stack, or only invoke the restart under special conditions. This works as the \code{lambda} of the \code{handler-bind} is invoked at the point of error on top of the stack, where the \code{continue} restart is visible.

Restarts and handlers are fantastic tools to create a more error friendly environment and to recover from unfavourable situations. Combined with the dynamic debugger and runtime compilation environment, these tools make debugging problems and evolving a system over time a lot easier.

@@

Additionally, Common Lisp offers several compiler tuning declarations. For instance, increasing the \code{speed} switch to its maximum will tell the compiler to try and optimise the code as much as it can. SBCL will then also emit warnings about cases where it doesn't have sufficient information to eliminate dispatch, or where it must allocate on the heap. Using this extra information the programmer can then refactor the code or add additional promises to ensure the compiler can generate efficient code.

Another switch is the \code{safety} switch, which when set to its minimum will tell the compiler to avoid inserting safety checks such as runtime type checks, bounds checks, or other features that typically help catch bugs. Naturally this is a dangerous option, as the generated code now becomes as unsafe as C or assembly code, and misuse of functions declared with this option can lead to memory corruptions.

@@

Finally you can also declare functions to be \code{inline}, allowing the compiler to inline the definitions at call sites. This comes with the usual advantage of aiding inference and removing call overhead, but does hamper dynamism, as a change to an inlined function now requires recompiling call sites as well.

@@

These techniques mostly focus on type dispatch within a function, with special focus on arithmetic functions. Dispatch for generic functions cannot be eliminated like this, and especially generic functions with a lot of methods attached or complicated method signatures can become a significant choke point.

This can be largely combated with traditional techniques of simplifying method signatures, splitting up generic functions, or rewriting them into statically dispatching functions that can be inlined to take advantage of type inference. We are also investigating an alternate dispatch technique outlined by Robert Strandh\cite{strandh2014fast} that should significantly improve generic function dispatch time, potentially even outperforming traditional vtable based approaches as used in C++ or Java.

@@

For instance, we can ask the compiler to allocate a number of objects on the stack by using a declaration (\code{(declare (dynamic-extent x))}). However, this only works if the compiler can know ahead of time what size the object is, meaning only fixed size arrays and known structs can be stack allocated. Furthermore, stack space is much more limited, so bigger allocations still have to happen on the heap.

For cases where the object is a class instance, simply too big for the stack, or has to escape the stack sometimes, we can still easily cache the instance locally by using \code{load-time-value}.\\\code{load-time-value} is a special operator that causes the form it wraps to be executed when the code is first loaded into the system, and then puts the result of that executed form in the place in the code the \code{load-time-value} was at. This effectively creates an anonymous global variable where the object is stored, which is then put in place of the original \code{load-time-value}.

@@

Another point where \code{load-time-value} becomes very useful is when objects are said to be immutable, and their construction arguments are known at compile time. In this case, we can define a ``compiler macro'' which is a macro that is executed to replace a call to a particular function. This compiler macro can then analyse the arguments to the function at compile-time and choose to substitute the call with a \code{load-time-value} form if all arguments can be determined statically. This then avoids the construction and allocation of the object, or even the entire computation of the function at runtime in a, to the user, completely transparent fashion.

@@

\section{Conclusion}
The expressiveness of the Common Lisp Object System allows for rapid changes in game object behaviour, as well as the creation of reusable pieces of behaviour that can often be seamlessly combined together. With extensions to the Meta Object Protocol, this can even be extended to graphics rendering routines and shaders.

The presence of the entire compiler environment at runtime allows the developer to change any part of the game while it is running. This is especially feasible thanks to the presence of the restarts and interactive debugger, which allow the game to recover from a bug without having to crash and restart the entire system.

@@

\section{Acknowledgements}
We would like to thank the very handsome and attractive people that helped proofread this draft ;)

>>>>

We explore in detail some of the aspects of Common Lisp that make it particularly suited for games, and also discuss some of the pitfalls we encountered and how to combat them.

@@

In CLOS, methods are not attached to classes, but are instead parts of a generic function. Methods ``specialise'' to one or several of the required arguments of the function to a class. When the generic function is called, the system must first determine the set of ``applicable methods''. This set depends on the method combination used by the generic function. For brevity's sake, we will only look at the standard method combination here.

@@

As a brief example, let us consider \autoref{lst:method definition}. Here we define a generic function of two arguments called \code{handle}, as well as three methods. Each of the methods requires the first argument to be a subclass of \code{tick}. Method 1 and 2 also require the second argument to be a subclass of \code{player}, and method 3 a subclass of \code{enemy}.

@@

This method combination mechanism really shines only once we consider inheritance and especially multiple inheritance and the arising ``mixin classes''. Let us now define the classes we used for the previous listing. \autoref{lst:inheritance} shows five classes being defined, with \code{tick} being a subclass of \code{event}, and \code{player} and \code{enemy} being subclasses of \\\code{physics-object}.

@@

In other OOP paradigms a similar behaviour can usually be achieved by calling the method of the superclass with \code{super}, but notice here that the behaviour of method 2 does not require changing anything about the other methods or subclasses. It thus remains wholly encapsulated in its own behaviour.

So far so good. Now imagine that we decide enemies should emit a light so that they're always visible. To implement this behaviour, we're going to define a new class, \code{emitter}, and define the light flicker behaviour in a new \code{handle} method as shown in \autoref{lst:emitter}.

@@

To make the enemy adopt this new behaviour, all we need to do is add \code{emitter} to the \code{enemy} list of superclasses of the enemy class. Thanks to the method combination, calls to \code{handle} will now include the \code{emitter}'s method, and we have achieved the combination of the two behaviours.

@@

The order of the superclasses here is important. Classes that appear earlier in the list have ``precedence'' and are thus considered to be ``more specific'' when the set of applicable methods is determined. By having \code{emitter} appear after \code{physics-object}, we ensure that the \code{emitter}'s \code{:before} method runs \textit{after} that of \code{physics-object}, ensuring the collisions have already resolved properly once we consider the lighting update.

@@

One issue that crops up when behaviours are segregated into such small compartments is that you might not have enough control over the combination of them. The standard method combination offers no fine-grained control to exclude or reorder methods for a specific specialisation of an argument. One could devise their own method combination strategy to allow this kind of specialised behaviour, however we are not convinced that this would not lead to an ultimately even more confusing design.

While the standard method combination and mixins offer a great deal of flexibility out of the box, great care must still be taken when the overall class hierarchy and function protocol is designed. Otherwise behaviours will not combine cleanly and lead to strange bugs, or might not combine properly at all.

@@

The first part is what's called ``conditions'', or Exceptions in other languages. Conditions are ``signalled'' (thrown) and can then be ``handled'' (caught) by a piece of code lower down in the stack. When handled by a standard condition handler, the stack unwinds to the point of the handler, and the matching handler function is invoked to resolve the issue. For our purposes here, Common Lisp offers two types of conditions: warnings and errors. When a warning is signalled that is not handled further up the stack, it simply vanishes and execution continues from the signalling point unimpeded. If an error remains unhandled, a dynamic debugger is invoked instead and the signalling thread is paused. From there one of the unique features comes to shine.

@@

This means that unlike traditional languages where an error causes a stack trace and then tries to continue or simply crashes, in Common Lisp execution is halted, waiting for you to analyse and potentially fix the bug. Once you're confident you know what to do or have solved the issue, you can pick from one of several ways to continue the execution from the error. Doing so leaves the program running at all times and means you don't need to worry about long recompilation times or setup times to reproduce the error. You can immediately continue where you left off.

@@

The final piece is the existence of \code{handler-bind}. This is a type of handler for conditions, but unlike the regular handlers that cause the stack to unwind, these handlers are invoked \textit{on top of the stack} of where the condition was signalled. As such, they have full access to the dynamic environment at the point of the error. In fact, this is also how the top level debugger is invoked. However, with \code{handler-bind} you can automate the error resolution and avoid requiring user input.

A system with well designed recovery points through restarts can then be set up to automatically recover from most errors in a way much more refined than traditional Exceptions, as there is an inversion of control.

@@

However, importantly, for the \code{main} \code{render}\\ method, this restart is automatically invoked, ensuring that any potential crashes at runtime of the game are gracefully ignored. This handler could also be defined elsewhere even higher up the stack, or invoke the restart only under special conditions. This works as the \code{lambda} of the \code{handler-bind} is invoked at the point of error on top of the stack, where the \code{continue} restart is visible.

Restarts and handlers are fantastic tools to create a more error-friendly environment and to recover from unfavourable situations. Combined with the dynamic debugger and runtime compilation environment, these tools make debugging problems and evolving a system over time a lot easier.

@@

Additionally, Common Lisp offers several compiler tuning declarations. For instance, increasing the \code{speed} switch to its maximum will tell the compiler to try and optimise the code as much as it can. SBCL will then also emit warnings about cases where it doesn't have sufficient information to eliminate dispatch, or where it must allocate on the heap. Using this extra information, the programmer can then refactor the code or add additional promises to ensure the compiler can generate efficient code.

Another switch is the \code{safety} switch, which, when set to its minimum, will tell the compiler to avoid inserting safety checks such as runtime type checks, bounds checks, or other features that typically help catch bugs. Naturally this is a dangerous option, as the generated code now becomes as unsafe as C or assembly code, and misuse of functions declared with this option can lead to memory corruptions.

@@

Finally you can also declare functions to be \code{inline}, allowing the compiler to inline the definitions at call sites. This declaration comes with the usual advantage of aiding inference and removing call overhead, but does hamper dynamism, as a change to an inlined function now requires recompiling call sites as well.

@@

These techniques focus mostly on type dispatch within a function, with special focus on arithmetic functions. Dispatch for generic functions cannot be eliminated like this, and especially generic functions with a lot of methods attached or complicated method signatures can become a significant choke point.

This can be combated largely with traditional techniques of simplifying method signatures, splitting up generic functions, or rewriting them into statically dispatching functions that can be inlined to take advantage of type inference. We are also investigating an alternate dispatch technique outlined by Robert Strandh\cite{strandh2014fast} that should significantly improve generic function dispatch time, potentially even outperforming traditional vtable based approaches as used in C++ or Java.

@@

For instance, we can ask the compiler to allocate a number of objects on the stack by using a declaration (\code{(declare (dynamic-extent x))}). However, this works only if the compiler can know ahead of time what size the object is, meaning only fixed size arrays and known structs can be stack allocated. Furthermore, stack space is much more limited, so bigger allocations still have to happen on the heap.

For cases where the object is a class instance, simply too big for the stack, or has to escape the stack sometimes, we can still easily cache the instance locally by using \code{load-time-value}.\\\code{load-time-value} is a special operator that causes the form it wraps to be executed when the code is first loaded into the system, and then puts the result of that executed form in the place in the code the \code{load-time-value} was at. Another way to understand it is that an anonymous global variable is created where the object is stored, which is then put in place of the original \code{load-time-value}.

@@

Another point where \code{load-time-value} becomes very useful is when objects are said to be immutable, and their construction arguments are known at compile time. In this case, we can define a ``compiler macro'' which is a macro that is executed to replace a call to a particular function. This compiler macro can then analyse the arguments to the function at compile-time and choose to replace the call with a \code{load-time-value} form if all arguments can be determined statically. This substitution then avoids the construction and allocation of the object, or even the entire computation of the function at runtime in a, to the user, completely transparent fashion.

@@

\section{Conclusion}
The expressiveness of the Common Lisp Object System allows for rapid changes in game object behaviour, as well as the creation of reusable pieces of behaviour that can often be seamlessly combined together. With extensions to the Meta Object Protocol, this flexibility can even be extended to graphics rendering routines and shaders.

The presence of the entire compiler environment at runtime allows the developer to change any part of the game while it is running. Doing so is especially feasible thanks to the presence of the restarts and interactive debugger, which allow the game to recover from a bug without having to crash and restart the entire system.

@@

\section{Acknowledgements}
We would like to thank Robert Strandh, Florian Hanke, Tobias Mansfield-Williams, and Selwyn Simsek for their feedback on the paper.
